PanGraph: scalable bacterial pan-genome graph construction

### 2023-03-08 reviewer_response.tex

      There was no comparison with other tools - would be appropriate if e.g. downstream mapping performance were evaluated but that is outside the scope here.\\

      On reading I had a few queries, which could perhaps be clarified further in the text.}

\response{We thank the reviewer for their appreciation of our contribution and for their critical review of our work.}

\reviewer{R1.1}{A pancontig is defined as a multiple sequence (sub)alignment. In line 54 you say ``Pairwise graph alignment is performed by an all-to-all alignment of the pancontigs'' which I interpreted to mean all sequences in the alignment vs all sequences in another alignment. You do later add that ``To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned''. Adding the word consensus in the earlier sentence might avoid confusion.}
% \Marco{changed to ``Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree." Too compressed? I could also break in two sentences and expand a bit more.}\\
% \Liam{that's clear to me.}
\response{We agree with the reviewer's suggestion. The wording has been updated to ``all-to-all alignment of the consensus sequences of all \textit{pancontigs}''.}

@@

\reviewer{R1.9}{Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.}
% \Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}\\
% \Liam{think just adding a comment in the SI saying this is required.}
\response{Homologous recombination is not directly included in our simulated data generation procedure. We only simulate transfer of sequence between two genomes but this transfer is not based on homology. Moreover we simulate sequence evolution without a notion of genes. This is now specified in SI section \SIsynthdata. To avoid confusion we also changed ``Horizontal Gene Transfer'' to ``Horizontal Sequence Transfer''.}

\reviewer{R1.10}{Supplementary Information C: You comment that the use of null energy parameters results on average in ``more and shorter pancontigs'' but also that they result in more diverged sequences being merged into the same pancontig. Why is a more diverse pancontig more likely to become fragmented?}
% \Marco{I think that basically the reason is that merging more distant paralogs/orthologs can fragment more the graph. Maybe we could explain this more explicitly in the text.}\\
% \Liam{makes sense and explaining this reasoning sounds good.}
\response{We use the pseudo-energy defined in eq.~(1) as an heuristic to decide whether a homologous match detected by our aligner should result in a merge between two pancontigs, depending on the match length and on sequence divergence. In particular any choice of $\beta$ would prevent mergers between sequences with divergence higher than $1/\beta$. Setting $\beta=0$ instead does not impose a divergence limit on mergers other than the one implicitly set by the aligner sensitivity. In this sense the choice of null energy parameter results in more diverged sequences being merged into the same pancontig. A consequence of more merging is also more fragmentation of the graph, since the merging of a homologous section of two pancontigs often requires splitting them in chunks. Therefore we expected that the choice of null energy parameters would lead to shorter pancontigs, and alignments containing more diverged sequences.
      To make this reasoning more explicit we added section {\SIalgo} in the SI to better explain the merging procedure and the effect of tuning the pseudo-energy parameters.}

@@

      Nevertheless, there are a number of clarifications which need to be addressed before the paper is suitable for publication.}

\response{We thank the reviewer for this succinct summary of the field and the essence of our contribution.}

\subsubsection*{Major Issues}

\reviewer{R2.1}{Line 46: when you say ``PanGraph transforms an arbitrary set of genomes into a graph that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms.'' - I am a bit unclear on how the nucleotide level variation is summarised. The paper is a bit hand-wavey about this. When i dug around on github , i found this helpful page \url{https://neherlab.github.io/pangraph/tutorials/tutorial_2/} Which answers and raises questions. So, my question: for each pancontig (which I think is the same as a block), I think what you do is infer a consensus (what is this? majority base at each position but somehow also choose an allele when there is an indel?), and then you describe the other alleles in the MSA as edits from this? You surely do not do all pairwise descriptions of SNPS, right? On that same webpage you have a section on ``block consensus to node alignment'' - how do you choose the order of alleles here? This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.}
% \Marco{this point was also raised by R3. Mutations are always represented as variation on the consensus. The way consensus is propagated during graph merging might require more explanation. Maybe we could add a section in the supplementary with a somewhat more detailed description of the algorithmic steps we perform when merging two graphs? In the code \href{https://github.com/neherlab/pangraph/blob/e1c11012a3d19975d4b1dc61b1af3eb3949f7cd2/src/block.jl\#L1274}{this part} takes care of merging alignments between two blocks with only small (i.e. below cutoff) insertions and deletions. In short what we do is stitch the two alignments together using the cigar string as a guide, and re-compute the consensus on the newly-obtained alignment. This ensures that the consensus agrees with the majority of sequences in the alignment, but does not solve the full multiple sequence alignment problem. However we also provide the ``polish'' step for this, that can be executed after graph creation. This will take care of refining the block alignments.}\\
% \Liam{I think a `narrative section' in the supplementary describing this in brief sounds great.}
\response{This point was also raised by other reviewers and we appreciate the need to make this point clearer.
      In our pangenome graph representation each pancontig encodes an alignment using a consensus sequence and list of edits required to obtain each individual sequence from the consensus sequence.
      These edits consist of single nucleotide changes (mutations) as well as insertions and deletions.
      The consensus sequence is defined using the majority rule on the alignment and is updated after each merger.

@@

\begin{figure}[htb]
      \includegraphics[width=.7\textwidth]{figs_response/hp_vs_kp.pdf}
      \caption{frequency and length distribution (respectively left and right panels) of pancontigs for the \textit{Helicobacter pylori} and \textit{Klebsiella pneumoniae} datasets (top and bottom). For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter values and asm20 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange). Distributions are weighed by pancontig length, so that the sum of all bins heights is equal to the total pangenome graph size.}
      \label{fig:kp-vs-hp}
\end{figure}

\reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be ``left out'' in between pancontigs?}
% \Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\
% \Liam{if they possibly misunderstood, suggests that some statement that all sequence ends up in a pancontig should be added to manuscript. I suspect they are thinking of case of short ($<$100bp) stretches `in between' pancontigs, which (if I understand correctly) will be incorporated as insertions at the end of the consensus alignment of the adjacent pancontig.}
\response{No, every part of the sequence of each genome must be included in some pancontig. It might not necessarily appear in the pancontig consensus sequence if it is present in the alignment as a short ($< \Lthr$) insertion, but it will nonetheless be present in the alignment. In the particular case in which merging two pancontigs would leave overhangs shorter than $\Lthr$, these are included in the alignment as flanking insertions. This is now explained in SI section {\SIgraph}, and mentioned in the main text in section IB (line 125).
In fact, we used the fact that genomes can be reconstructed exactly from the pangraph to test and debug the tool. }

@@

\reviewer{R2.3}{line 48: ``Pancontigs are connected by an edge if they are syntenic on at least one input sequence;'' - i think this is a bit of an abuse of modern language. The word syntenic is generally used to mean two genes/whatever have order conserved on multiple chromosomes (although some googling tells me that in classical genetics it meant things were on the same chromosome). Anyway, my understanding is pancontigs are connected by an edge if one of the input genomes is spelled out by going from one to the other. So it's about adjacency rather than synteny.}
% \Marco{I agree, we could use ``if they are ajacent/contiguous on at least one input sequence''?}\\
% \Liam{adjacent sounds good.}
\response{We agree with the reviewer's observation. We have changed ``syntenic on'' to ``adjacent in''.}

\reviewer{R2.4}{Line 131: was N really=100? on the plot in figure 2, you have n.isolates going up to 1000}
% \Marco{$N=100$ was the standard value for the rest of the simulations (e.g. later for accuracy). I did a tentative correction: removed from the main the specification on population size and mentioned to check the SI for details on standard parameter values.}
\response{The value $N=100$ is the standard value used in the rest of the simulations (e.g. fig.~3), but as the reviewer pointed out this value is varied for the results showed in fig.~2 and the statement was confusing. We modified it removing the value of $N$ and referring to SI table~I for the standard value of simulation parameters.}

\reviewer{R2.5}{Figure 5 legend, lower case C in ``coli'' of ``E. coli''. Also italics.}
\response{The legend has been corrected.}

@@

\subsection*{Reviewer 3}

\reviewer{R3}{The article ``PanGraph: scalable bacterial pan-genome graph construction'' reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure. Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose. The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.\\
      For instance the authors state: ``Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.''. There are many pan-genome analysis tools which treat the pan-genome as a graph and generate a PGG. There are two primary differences in these existing tools: those which generate a graph at the nucleotide level with nodes representing basepairs (sometimes nonbranching runs of nodes are collapsed) and those where nodes represent larger features such as genes.\\
      PanGraph appears to wed these approaches to some extent in that the alignments are done at the nucleotide level and the features are extracted as a characteristic of the alignments.}

\response{
      We apologize if our previous discussion of the literature was incomplete. It was not our intention to downplay prior work.
      We thank the reviewer for highlighting additional important work and aspects that we have now added to better contextualise our work.
      We have rewritten the introduction to reflect previous work in this field and added to the discussion to further highlight previous and alternative approaches. We hope that these edits address the concerns of the reviewer.
      % Liam: I think good to say in more detail what has been added here. 
}
\Marco{Tentative extension - to be updated depending on how we modify the intro/discussion:\\

To amend this we performed the following changes:
\begin{itemize}
    \item The sentence in the abstract mentioned by the reviewer has been removed and modified.

    \item We extended the discussion of other methods in the introduction.\\
    In particular we extended the paragraph on methods based on de Bruijn graphs, mentioning also compacted generalization and the fact that tools exist that are able to build graphs containing thousands of isolates. We now cite more tools and put more emphasis on their applications, restricting it to tools specific for microbial genomics (e.g. we did not include PGGB). We removed the previous claim that homology detection was limited by the k-mer size, and cite tools that can be used to extend homology detection through small sequence modifications.\\
    We expanded in the same way the paragraph on homology-based and gene-based approaches, in which we now discuss more tools putting emphasis on their applications.

    \item We added a paragraph in the discussion in which we tried to emphasize how \textit{PanGraph} compares to gene-based and k-mer based PGG methods.\\
    In essence, methods based on de Bruijn graph have the advantage of being fast since they do not require alignments. They also do not necessarily require fully assembled genomes. However in these models the graph encodes in the same way nucleotide-level variation and large-scale structural variation, and specific post-processing of the graph is needed to distinguish the two. Conversely, PanGraph requires fully-assembled genomes and is potentially slower, but its pangenome graph representation provides a natural separation between small sequence differences (mutations or small indels) and big structural changes (gene gain/loss, inversions, rearrangements).\\
    Gene-based pangenome graphs have instead the advantage of being smaller in size (a node corresponds to a gene) and they can include more divergent sequences, since homology detection can be performed on protein sequences. Differently from homology-based methods like PanGraph however, they are often sensitive to annotation errors.
\end{itemize}
}

\reviewer{R3}{An obvious lack in PanGraph is an attempt to map existing annotations back onto the generated PGG which is a necessity for usefulness to biologists. The mechanism for this is obvious but providing a useful user interface might be tedious.}
% \Liam{I think perhaps producing a helper python script that converts a genbank annotation file for a genome within a pangraph to an output gff/csv with the block locations and coordinates would please this reviewer.}\\
% \Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
\response{We thank the reviewer for this important suggestion and appreciate the need to include a way to map annotations back onto the pangenome graph. We created a helper script that allows a mapping of pancontig information onto a set of annotations in a \texttt{gff} file. In default usage the pancontig location(s) of each feature gets added as a new `attribute', allowing a quick connection between the generated pangenome graph and existing annotations. The script and instructions on its usage can be found at \url{https://github.com/liampshaw/pangraph\_annotations}. In the future, we hope to extend this further for visualization purposes.
      % TODO: mention this in the text?
      % Liam: not sure if we need last sentence
}

@@

In the future we plan on providing a dedicated command to perform a more robust form of paralog splitting, which combines both sequence divergence information and context information. This command would build a phylogenetic tree from the alignment of each duplicated pancontig to detect exceptionally long branches, and check whether the split correlates with a split in the duplication context. This functionality is not however not yet implemented in the current version of \textit{PanGraph} and we see it as substantial extra work.\\

We appreciate the reviewer's perspective and have added SI sect.~{\SIalgo}D to include discussion of these considerations.

% TODO: confirm with Nick the paralog split algorithm.
}

\reviewer{R3}{In the pseudo-energy discussion a further parameter is mentioned: ``In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.''. The value chosen for this parameter should be explicitly stated along with a rationale for the choice. It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.}
% \Liam{this relates to R2's point above about all sequence ending up in pancontigs so I think is just clarification.}\\
% \Marco{I agree. We set a minimal cutoff of 100bp. This is stated later in line 106. Maybe we could expand the sentence in line 99 and give the value and few more words on the meaning/effect. This controls in some sense the ``corse graining'' of the graph: what is the size of an event that we want to encode as structural variation in the graph, rather than as an indel in the alignment. In the algorithm this minimal cutoff has different effects depending on the context. When aligning, if the match lenght is less than this threshold the merge is refused. For long alignments, if the CIGAR string contains indels that are longer than this threshold, these are saved as separate blocks and not as indels in the block. The same goes for the flanking sequences, if merging two long blocks leaves overhangs that are less than this threshold size then these overhangs are saved as part of the merged block and not as separate small blocks. This might be explained in the SI?}\\
\response{The minimal size parameter, now named $\Lthr$, is set to 100 bp in PanGraph. This parameter controls the minimal size of pancontigs. In general, this parameter sets a scale separation between structural variation that is encoded in the graph structure or in the pancontig alignment. For instance, an insertion could appear either as a separate pancontig, or as an insertion in the pancontig alignment, depending on its size. In practice this parameter is used in three ways:
      \begin{itemize}
            \item In addition to the pseudo-energy costraint, merges are refused if they have alignment length shorter than $\Lthr$.
            \item During merging, if the alignment contains long ($> \Lthr$) indels, these will cause a split in the merged pancontigs and will be treated as separate pancontigs.
            \item If a merger of two pancontigs would produce flanking overhangs shorter than $\Lthr$, these are not saved as separate pancontigs but are included in the alignment as flanking indels.
      \end{itemize}
      We choose a value of 100 bp in our analysis (and as standard value) to be able to detect variations at the gene scale, while encoding variations of up to few tens of bps in the alignments. This value can however be tuned by the user depending on the specific needs.\\
      We appreciate the need to better explain the role of this parameter in the merging procedure and in the structure of the resulting graph. To this end we extended its definition in section IB in the main text, and give a more extended discussion in SI section~\SIalgo B and \SIalgo C.\\
      % TODO: Homogenization in the main text with discussion on precision. Extend.
}

\reviewer{R3}{How are pancontig consensus sequences determined and updated? I could not find any details about this. This is not a trivial operation. My reading of the article seems to imply that a complete multiple sequence alignment is not done for each merger. Does the consensus sequence degenerate iteratively with mergers? For the simulated data how accurate are the final pancontig consensus sequences?}
% \Liam{expanding discussion of doing an MSA at end as for R2's comments would partially address. Quantifying how the consensus sequence degenerates iteratively is an interesting point that seems possible?}\\
% \Marco{I agree, this could be addressed by adding a more detailed algorithm explanation of the merging algorithm in the SI. The short answer is that at each merging the full alignment is constructed using the CIGAR of the alignment of the consensus of the two blocks. Then consensus of the full alignment is re-evaluated. This does not solve the multiple alignment problem during constructions and there can be degeneration on where indels are mapped, but the final graph can be eventually polished. I don't think we ever quantified how much the consensus sequence degenerates during the iterations. Maybe one could devise a proper test set and see how it degenerates over the iterations.}\\
\response{This point was raised by other reviewers, and we appreciate the need to make this part of the alorithm clearer. To this end we added section {\SIalgo}B in the SI, with a detailed explanation of how the merging procedure works.\\
      In essence, during graph merging alignment is only performed between the consensus sequences of the merged pancontigs. We avoid performing a multiple sequence alignment during merging to maintain the algorithm scalable. However we do update the consensus of the pancontig at each merge using the majority rule on the merged alignment.\\
      The idea behind this approach is that the consensus sequence is needed to find large-scale homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an exccessive amount of resources. However after the graph has been built alignments can be refined with \textit{PanGraph}'s \texttt{polish} command. This command performs a multiple sequence alignment on all the sequences in each pancontig and updates the consensus sequence accordingly. This separates efficiently the problem of resolving the large-scale and nucleotide-level diversity between sequences.\\
      We did not quantify explicitly the divergence of consensus sequences during graph building, but we observed that, even without ``polishing'', alignments of up to hundreds of sequences look consistent.
      % TODO: quantify this in simulations if there is time?
      % Liam: Think the last sentence is potentially risky - I don't get what `consistent' means? My rewrite (not confident in this): In our experience, the alignments appear reliable throughout this process, although we have not explicitly quantified it from our simulations. However, for cases where the alignments are of interest we recommend running `polish' to perform MSA for pancontig consensus sequences (as in the documentation) to ensure that the final alignments are as accurate as possible.  
}

### 2023-03-08 reviewer_response.tex

\reviewer{R3}{The statement: ``Upon completion, transitive 108 edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent pancontigs.'' needs clarification. Why would these pancontigs have been split in the first place? Is there any necessary special issues with producing the merged multiple sequence alignment?}
% \Liam{don't understand their last sentence. Maybe should quantify how common the transitive edges typically are?}\\
% \Marco{I can think of one edge case which is due to periodic boundary conditions. Initially each circular genomes is saved as a single long block. If during alignment this unique blocks is split somewhere in the middle, the resulting edge blocks might be always colinear, since they are actually adjacent parts of the sequence that were artificially split to linearize the circular genome. One wants to collaps these two parts in a single block. This should happen often at the beginning of the merging of circular genomes. But might be unnecessary detail when explaining the algorithm the first time, we might want to move it to the SI section where we explain merging in more detail?}\\
\response{Transitive pancontigs are a small but problematic artifact of periodic boundary conditions when merging circular genomes, such as bacterial chromosomes. At the beginning of the procedure the circular genome is saved as a single linear pancontig. This introduces an artificial cut. When during merging this unique pancontig is further split, the flanking pancontigs could be crossing the artificial cut in a co-linear way on all paths. This step of the procedure resolves the artificial cut by merging these pancontigs together.\\
      We think that the mention of this step of the algorithm in this part of the paper might be a bit confusing and needs more clarification. We moved and expanded the statement from section IB of the main text to SI section {\SIalgo} (point 5 of the merging algorithm explanation).}

@@

% To amend this we added the size of soft-core genome for each species in table~1 in the main text, and added part of this discussion in SI sect.~{\SIdataBenchmark}A.
% TODO: core genome for simulated data if there is time?

%     "all core": 778,
%     "all softcore": 2081,
%     "all genes": 23128,
%     "all nondupl": 18798,
%     "drop core": 822,
%     "drop softcore": 2069,
%     "drop genes": 23054,
%     "drop nondupl": 18749

\reviewer{R3}{The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.}
% \Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
\response{
      We appreciate the importance of the point raised by the reviewer.
      We agree with the reviewer that this post-processing smoothing of the graph produced by of \textit{PanGraph} would be a valuable addition, and we hope to include such further post-processing refinements in subsequent versions. 
      This sort of refinement would allow alignments of adjacent edges of pancontigs to be harmonized, potentially also removing very small pancontigs when these could be appropriately absorbed in the alignment of larger ones. We hope to continue to expand the downstream post-processing of \textit{PanGraph} data structures in future work. 
      % Liam: tried to make it sound more explicitly 'future' and 'further'
}

@@

\bibliography{cite}{}

\end{document}

%  TODO: before submission
% - [ ] archive repository state and add in the response
% - [ ] add files with accession numbers and pairs to the submission
% - [x] add mention of ``polish'' in main text.
% - [ ] for the future: add sequence reconstruction to pangraph automatic testing.
% - [x] check that all SI sections have been mentioned in the main text
% - [x] make Liam's repo public

### 2023-03-08 main.tex
Update on Overleaf.

We simulated populations of size $N$ and sequence length $L$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~{\SIsynthdata} for details on the simulation and standard parameter values).
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line tools for external use.

>>>>

We simulated populations of size $N$ and sequence length $L$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~{\SIsynthdata} and SI table~I for details on the simulation and standard parameter values).
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line tools for external use.
